---
title: "In class assignment"
subtitle: "University of Tennessee - AGNR220/SNR220"
format: 
  html:
    embed-resources: true
editor: visual
---

```{r}
#| include: false
knit_print.data.frame = function(x, ...) {
  res = paste(c("", "", 
                knitr::kable(x, digits = 2) |> 
                  kableExtra::kable_styling()), collapse = "\n")
  knitr::asis_output(res)
}

registerS3method(
  "knit_print", "data.frame", knit_print.data.frame,
  envir = asNamespace("knitr")
)

```

# Dairy Farm🐄

A dairy farm 🐮 in Wisconsin has purchased a new dietary supplement that claims to increase the milk production. The farm has 1,000 cows, and they are giving the supplement to half of them.

After a while, you decide to measure the amount of milk production in 40 cows taking the supplement, and in 40 cows that are not taking the supplement.

Please download the datasets from `Canvas`.

Read them into R and convert them into a vector. Look at the following code:

(You can copy the code)

```{r}
CowControl<-read.csv("CowControl.csv")
CowSupplement<-read.csv("CowSupplement.csv")
CowControl<-CowControl$x
CowSupplement<-CowSupplement$x
```

## Step 1.

Use the `sample()` function to sample 40 individuals from the `CowControl` object. Name this object `control`. Use the `sample()` function to sample 40 individuals from the `CowTreatment` object. Name this object `treatment`.

```{r echo=F}

control<-sample(CowControl,40)
supplement<-sample(CowSupplement,40)
```

Then, copy and paste the following code, to run your t test! Easy! 😃

```{r eval=FALSE}
t.test(control,supplement, var.equal = T)
```

## Assignment Question 1 🐮

Based on your test, make a statistical decision. Does the supplement have an effect?

## Assignment Question 2 🐮

New information:

-   The supplement costs \$30.00 for 1,000 g.

-   Each cow needs to take 100 g. each day

-   The farm sells each liter of raw milk to a processing facility at \$0.40
