---
title: "Assignment GLM"
subtitle: "GLM (logistic)"
author: "Alejandro"
format: 
  html:
    self-contained: true
    toc: true
editor: visual
---

## Generalized Linear models

Just as a reminder... We use GLM's when the predictor and response variables **DO NOT** have an underlying normal distribution of the residuals or lack an underlying linear relationship.

In **Linear Models** we model the response as a function of the predictors.

**In GLM,** we model **a function** of the response as a function of the predictors. That function is called the link function. So, we are not modeling the response variable (usually "y") but a link function of the response. So, instead of modeling:

$$
\hat{y} = \beta_0+\beta_1x_1 
$$

(where $\hat{y}$ represents the **expected** value of y) we model:

$$ logit(p) = \beta_0+\beta_1x_1 $$or

$$ log(\lambda) = \beta_0+\beta_1x_1 $$

We use logit for logistic regression (aka binomial distribution), and we use log for Poisson distributed data.

::: callout-warning
## Think about it 🧠

How to interpret the $\beta's$? 🤔

Everything to the right of the `=` sign is **THE SAME.** This means, the link function allows us to estimate link(y) as a linear model!

In the simple linear model the $\beta_0$ is the expected (or predicted) value of y when x is 0. The $\beta_1$ is how much y changes for a one unit increase in x. This is called the slope and it is very useful! it can tell us how strong the effect of x is on y (effect size). We can analyze whether this effect is biologically important. We can analyze whether it is **significantly** different from 0. All in all, the $\beta's$ are super informative. And as the paper we read said, inference is about the $\beta's$!

What happens when we have generalized models? The $\beta's$ represent effects **NOT** on the response variable, but on the **link function** of the variable. So $\beta_1$ represents how much $log(\lambda)$ changes for a one unit increase in x. But it doesn't tell us how much $\lambda$ changes. In order to figure this out, we need to do an inverse link equation:

$\lambda = e^{\beta_0+\beta_1x_1}$

For this reason is important to:

1.  Know the link functions
2.  Know the **inverse** link functions
3.  Know these functions in R
:::

Here is a table with the link functions for both Binomial and Poisson distributions.

Save it somewhere important!

| Distribution | link name | link equation | inverse link eq | link function R | inverse link in R |
|----|----|----|----|----|----|
| Binomial | logit | $\mu = log(\frac{p}{1-p})$ | $p= \frac{e^\mu}{1+e^\mu}$ | `qlogis()` | `plogis()` |
| Poisson | log | $\mu = log(\lambda)$ | $\lambda = e^\mu$ | `log()` | `exp()` |

## This assignment

We went through the Poisson example in class. You can also download the presentation file (from Canvas) and access all of the code we used. During this assignment we will focus on the following:

1.  Logistic regression
2.  Questions where you are presented with data, and a research question and you have to analyze the data. You need to decide what test or model to use, and interpret the results.

## Logistic Regression

We use the logistic regression when we have binomial data. Remember, in binomial data the response variable is binary, our responses are limited to 0's and 1's. Which is which depends on you, but usually 1 is seen as a "success" or "positive".

Some examples:

1.  Presence/absence
2.  Alive/Dead
3.  Homozygous/Heterozygous
4.  Mature/non mature
5.  Male/Female
6.  pregnant / not pregnant
7.  Healthy / Disease

::: callout-warning
## Think about it 🧠

What is the response variable? 🤔

We usually obtain the **mean** of the response variable. In the grasslandexample, the response variable was **KgDMHA,** or Kilograms of Dry Matter per Hectare.

In the logistic regression, we are trying to find the **probability** of an event happening. Look at the examples before this box. If the binary outcomes are pregnant or not pregnant, the response variable is the probability of being pregnant.

If we were exploring whether cortisol has an effect on succesful pregnancies in mice, and the model was `glm(pregnancy~cortisol, data=data, family=binomial(link="logit")` then, what we are trying to estimate are 1) whether cortisol has an effect on the probability of being pregnant (inference) **and** 2) what is the probability of a mouse being pregnant given a specific cortisol measurement. This explains why it is not linear. Probabilities go from 0 to 1.
:::

The link equation is the "log of odds", also known as logit. This equation allows us to move from a system were the values go from 0 to 1 (probability) to one where we can theoretically go from -INF to INF.

#### Log of odds? What is that? And what are odds?

The logit link function is: $\mu = log(\frac{p}{1-p})$ or also known as "*the log of odds"*. This works because **any** probability can be converted to log odds by finding the **odds** and taking the log of that. [**James Jaccard**](https://books.google.com/books?id=tj1Wn5u9gSMC&printsec=frontcover#v=onepage&q&f=false) calls log odds  "counterintuitive and challenging to interpret". They are not as easy to interpret as the "log" we use in Poisson. However, I don't think you absolutely need to understand the transformation, you only need to understand why it is useful!

First off, the odds is $\frac{p}{1-p}$. It is simply dividing the probability of an event occurring divided by the probability of it not occurring. Let's look at some examples:

In a coin toss with 0.5 probability, the odds ratio is: $\frac{0.5}{1-0.5} = 1$. We can interpret this as the odds of getting a success (or head) is 1:1 (think 1 to 1, or 50-50)

Let's imagine the probability of getting a 5 after rolling one die. The probability of this event is $\frac{1}{6}$, so, the odds are:

```{r}
(1/6)/(5/6)
```

The odds are 0.2. This means you will see 0.2 successes for every failure (or one success for every 5 failures).

Finally, let's think of the odds of me (Alejandro) seeing a car accident during my daily drive to and from campus). I know this is biased, and one semester I will actually sample this for an exercise, but my very biased estimate says the probability is 0.8 (80% chance of seeing one). If this was true, then the odds would be:

```{r}
0.8/(1-0.8)

```

Four. Again, think of this as four to one. If I drove five times, I would see an accident 4 times, and "no accidents" one time.

OK, that was a lot of time spent on odds. And the reason I did that, is that my brain struggles thinking in odds rather than probabilities. Which is why we want to present the results in a probabilistic scale. However, this concept is important to understand how it is estimated.

Odds are always "positive". But the transformation need to potentially be from -inf to inf. Similar to what we do with counts, then we take the log of the odds.

Let's imagine the probability of winning the lottery is 0.000000000065789. Then, the log odds would be:

```{r}
log(6.5789e-11/(1-6.5789e-11))
```

-23.44. And if we had an event with high odds (like the accident one):

```{r}
log(0.8/(1-0.8))
```

Here you can see how values higher than 1 (p higher than 0.5) will be positive, and odds lower than 1 (p lower than 0.5) will be negative.

Actually, we can plot the relationship between p (probability) and logit(pi):

```{r}
pi = seq(0.0001,1-0.0001, by=0.0001)
plot(pi~qlogis(pi), type="l",xlab="logit(p)",ylab="p"); abline(v=0, lty="dashed"); abline(h=0.5, lty="dashed")

```

What I want you to take away from this whole section is the following:

::: callout-tip
## Take home message 🏠

1- logit(p) ranges from -Inf to +Inf as pi increases from 0 to 1 Take home message

2- logit(p) takes on a full range of values which allow the modeling algorithm to explore a full range of coefficient values in the systematic component of the model In other words, $\beta_0 + \beta_1x_1 + \beta_2x_2 + ... + \beta_mx_m$ can be unconstrained and can take values from -Inf to + Inf

3- logit(p) is `qlogis(p)` in r
:::

#### Let's actually do a regression

This dataset (parasitecod.csv) was obtained from the following book:

Zuur, A., Ieno, E. N., Walker, N., Saveliev, A. A. & Smith, G. M. *Mixed Effects Models and Extensions in Ecology with R*. (Springer New York, 2009).

It's a highly recommended book (I showed it to you during the first week!)

Load it into R and let's run a glm.

```{r}
#| warning: false
cod<-read.csv("parasitecod.csv")
library(dplyr)
cod<-cod%>%mutate(across(c(Year,Sex,Stage,Area),as.factor))
```

Before running the glm, I changed some categorical data to "factor" in R. This step is really important, because if you don't do it, it will take the data as continuous!

Now, let's run the model!

```{r}
codmodel1<-glm(Prevalence~Length+Year+Area,data=cod,family = binomial(link="logit"))
summary(codmodel1)
```

Remember that those coefficients (think of each coefficient as a $\beta$, so this model has 7 $\beta's$) are for the link function:

::: callout-important
## Assignment question 1 📝

Look at the summary. And before continuing make sure you understand it. If you don't, now is the time to raise your hand.

You need to calculate the expected probability (we'll call this $\pi$) of an individual being infected for each of the following two cases:

1.  An individual of length 50 from Area 1, in 1999
2.  An individual of length 50 from Area 3, in 2001

Remember that:

$$
log(\frac{\pi_i}{1-\pi_i}) = \beta_0 +  \beta_1x_{1,i} + \beta_2x_{2,i} + ... + \beta_mx_{m,i} 
$$

And we are trying to solve for $\pi$

Please do not use the function "predict". You can use algebra to solve this problem, or you can use the `qlogis` and `plogis()` functions.
:::

We can use the packages `car` and `emmeans` to run an ANOVA and look for differences among some of the explanatory variables

```{r echo=F}
#| warning: false
library(car)
library(emmeans)
library(ggplot2)
```

First we run:

```{r}
Anova(codmodel1)
```

And see that there are significant effects of all three factors.

Let's explore the effects of Year running pairwise comparisons using `emmeans`:

```{r}
emmeans(codmodel1, ~Year) %>% contrast("pairwise")
```

Please note that 1) results are averaged over the levels of Area, and 2) results are given on the log odds ratio scale. This is **not** the response scale. The response scale is probabilistic, so it goes from 0 to 1 (and therefore there could not be a difference of 1.247

::: callout-important
## Assignment question 2 📝

Run a pairwise comparison for Area, and a pairwise comparison for Year and Area
:::

Finally, let's plot it!

First, let's predict the values. Predict also gives us values on the log odds scale, so, some transformation is needed:

```{r}
predictedmodel<-predict.glm(codmodel1,cod,se.fit = T)
ci_lwr <- with(predictedmodel, plogis(fit - 1.96*se.fit))
ci_upr <- with(predictedmodel, plogis(fit + 1.96*se.fit))
cod2<-cbind(cod,predictedmodel)
cod2$fit2<-plogis(cod2$fit)
cod2$lwr<-ci_lwr
cod2$upr<-ci_upr
```

And finally, let's plot it!

These type of models can be pretty tough to plot. One option is to use a grid, in which each column is a year, and each row is an area:

```{r warning=F, fig.height=5}
ggplot(cod2,aes(x=Length,y=Prevalence,ymin=lwr,ymax=upr))+
  geom_point()+
  geom_line(aes(y=fit2))+
  geom_ribbon(alpha=0.15)+
  xlab("Length (cm)")+
  ylab("Prevalence")+
  theme_bw()+
  facet_grid(Area~Year)
```

While tough to interpret, we can see some patterns. For example, fish from 2001 have a lower prevalence probability!

Another way to plot it, is to choose one categorical variable to be presented as colors, and the other one to be at the grid level:

```{r warning=F, fig.height=5}
ggplot(cod2,aes(x=Length,y=Prevalence,ymin=lwr,ymax=upr,color=Area,shape=Area,fill=Area))+
  geom_point()+
  geom_line(aes(y=fit2))+
  geom_ribbon(alpha=0.1)+
  xlab("Length (cm)")+
  ylab("Prevalence")+
  theme_bw()+
  facet_grid(~Year)
```

In here, we can see two patterns: 2001 has lower probability of prevalence, while area 4 has a higher probability. Finally, size also has an effect.

::: callout-important
## Assignment question 3📝

Look at the cod data and come up with 3 "biological hypotheses" that you can run as models. Run the three models

Use model selection or any method that you want to compare the 4 models (the model I ran; codmodel1; and the models you ran). And select the best model.
:::

::: callout-important
## Assignment question 4 📝

For the best model from question 2 (if the best model is the one I ran, then select the second best model) please do the following:

1.  Interpret the summary output
2.  Run an ANOVA (and if needed) a pairwise comparison
3.  plot the model. If the model seems too complex to plot, ask for my help, and we can figure it out together 😃
:::

## Testing your knowledge

Moving forward, for each assignment I will ask you 1 or 2 questions in which you will have to apply knowledge from previous classes and assignments.

::: callout-important
## Assignment question 5 📝

**Data:** `welldata.csv`\
1. wellID - ID of well\
2. fluoride - mg/L of fluoride in a sample\

**Research question:**

You are tasked with researching whether the mean content of fluoride in a large rural area with \>2,000 private wells might be over the EPA recommendation of 4.0 mg/L. You sample 28 wells.
:::

::: callout-important
## Assignment question 6 📝

**Data:** `parasitecod.csv`\
1. Intensity: Number of parasites present\
2. Prevalence: 1-parasite present, 0- no parasite\
3. Year\
4. Depth (in meters)\
5. Weight (g)\
6.Length (cm)\
7. Sex\
8. Stage\
9. Age\
10. Area\

**Be aware!** The weight, length, stage and age covariates are highly co-linear! Use only one at a time.

Explore 3 different hypotheses (make sure to include a null model as part of your hypotheses), but use intensity as your response variable.

**Be aware!** The `intensity` response variable is missing some data! Thee are cases were the number of parasites were not counted. You need to deal with this before you can run the models.
:::
