---
title: "Exam question"
subtitle: "Central limit theorem, CI's, T-test"
author: "Alejandro Molina-Moctezuma"
format: 
  html:
    self-contained: true
execute: 
  eval: false
editor: visual
---

# This lab

We are going to do a little experiment regarding the Central Limit Theorem and confidence intervals. We are going to:

1.  Simulate a population.
2.  Take a sample of size n = 40
3.  Estimate the mean, and CI of the sample
4.  Compare it to the population. Did we get it right? Is the real mean encompassed in the population?
5.  Do it again... and again... and again
6.  Do it 1000 times
7.  Plot this using ggplot
8.  Make some decisions.

This may seem like a lot... but it is not! It can be done fairly easily using R. And I will give you all the code!

EACH ONE OF YOU WILL GET DIFFERENT RESUKTS, AS WE ARE SIMULATING DATA

#### Step 1: Simulate your data

Go to file \> new file \> R script and create a new script.

Choose **only one** of the following examples! Copy the code for your example and run it once in the script

##### Example 1: bats 🦇

This data-set is on the wing aspect ratio of the pygmy round-eared bat.

```{r}
data<-rnorm(10000, mean = 5.5, sd=0.75) #your population, 10,000 individuals total
mu<-mean(data) #real population parameter
sigma<-sd(data)
mu
sigma
hist(data)
n<-40
```

##### Example 2: frogs 🐸

This data-set is on the presence-absence data of Chytridiomycosis in a population of [*Lithobates clamitans*](https://en.wikipedia.org/wiki/Lithobates_clamitans)This is a binomial distribution. 1 is presence and 0 is absence

```{r}
data<-rbinom(1000,1,0.23) #your population, 10,000 individuals tota. 
mu<-mean(data) #real population parameter
sigma<-sd(data)
mu
sigma
hist(data)
n<-40
```

##### Example 3: grapes 🍇

This dataset is on the number of wild grapes in 10,000 vines in a meadow (you can also make this wine grapes in a vineyard if you are into that).

```{r}
data<-rpois(1000,40) #your population, 10,000 individuals total. 
mu<-mean(data) #real population parameter
sigma<-sd(data)
mu
sigma
hist(data)
n<-40
```

##### Example 4: Beef 🐮

Annual live weight (LW) meat production in 10,000 hectares (n = 10,000), each hectare represents one datapoint.

```{r}
data<-rnorm(10000, mean = 410.5, sd=60) #your population, 10,000 individuals total
mu<-mean(data) #real population parameter
sigma<-sd(data)
mu
sigma
hist(data)
n<-40
```

### Step 2: Write it down

In your word document, write the mean for your population, write the sd for your population and paste the histogram. Answer: What does this histogram represent?

#### Step 3: Take a sample

This is what we are going to do: we are going to take a sample of size 40, obtain the mean, and the standard error. Then, we will take a new sample of 40, estimate the mean and standard error. And the again. We will do this 100 times. No worries, I will give you all the code.

CREATE A NEW SCRIPT:

Go to file \> new file \> R script and create a new script.

And in the new script, do the following:

1- Take the sample

```{r}
n<-40
```

```{r}

library(ggplot2)
experiment<-1:100
means<-rep(NA,100) #For the means
s<-rep(NA,100) #sd
SE<-rep(NA,100) #SE
LCI<-rep(NA,100) # Lower CI
UCI<-rep(NA,100) # Upper CI
inclusive<-rep(NA,100) #is the real parameter included?
alpha<-0.05 #our alpha
lower<-qnorm(alpha/2) #lower critical value
upper<-qnorm(1-alpha/2) #upper critical value
for(i in 1:100){
sample<-sample(data,n)
means[i]<-mean(sample)
s[i]<-sd(sample)
SE[i]<-s[i]/sqrt(n)
LCI[i]<-means[i]+SE[i]*lower
UCI[i]<-means[i]+SE[i]*upper
if(mu<UCI[i] & mu> LCI[i]){
  inclusive[i]<-T} else{
    inclusive[i]<-F
  }
}

df<-data.frame(experiment,means,s,SE,LCI,UCI,inclusive)

```

#### Step 4: Obtain Histogram:

Now run the following

```{r}
hist(means)
```

And copy the new histogram in your word file. This is a histogram of the actual estimated means. Are the estimated means around the real mean? How does this compare to the population histogram?

## Step 5: Plot:

Run the following plot:

```{r}
ggplot(df, aes(x = experiment, y = means,colour=inclusive,ymin = LCI, ymax=UCI)) +
   geom_point() +
   geom_hline(yintercept = mu, color = "black", linetype = "longdash") +
   scale_y_continuous(name = "", limits = c(min(df$LCI) - min(df$LCI)/10 , max(df$UCI + max(df$UCI)/10))) +
   scale_x_continuous(limits = c(0, 100)) +
   theme_classic()+
  scale_color_manual(values=c("#ec280e", "#06b512"))+
  geom_errorbar(width = 0)
```

Copy this plot in your word document and paste it.

Answer: How many times did the Confidence Interval included the real value. This is a 95% confidence interval, so, it should include it 95% of the time!

## Multiple simulations

Run the code on step 3 followed by the plot in step 5 multiple times. Like 4 or 5 times. You should get a different result and plot each time. Write the maximum and minimum number of times in which the CI did not include the real mean. Answer: Does this make sense taking into account that we had a 95% CI?
