---
title: "Assumption Checking, Transformations & Intro to GLMs"
author: "PLSC 571 — Experimental Design"
format: html
editor: visual
---

```{r setup, include=FALSE}
# Load all required libraries up front
library(car)      # for Anova() Type II/III
library(ggplot2)  # for plotting
```

## Overview

This in-class coding activity has **three sections**. All code is provided — your
job is to **run the code, look at the output, and answer the discussion questions
on Canvas**.

You will **not** write any code today. Read the prompts, run each chunk in order,
and submit your answers through Canvas.

------------------------------------------------------------------------

# Section 1 — Assumption Checking (~7 min)

We will work with a crop-yield dataset. Four wheat varieties (*Alpha*,
*Beta*, *Gamma*, *Delta*) were grown at different nitrogen application
rates. The response variable is **yield (kg/ha)**.

### Load the data and fit a linear model

```{r load-data1}
# Read the yield dataset
yield_data <- read.csv("data/yield_transform_dataset.csv")

# Make variety a factor
yield_data$variety <- as.factor(yield_data$variety)

# Quick look at the data
str(yield_data)
summary(yield_data)
```

```{r fit-raw-model}
# Fit a linear model on the RAW (untransformed) yield
raw_model <- lm(yield_kgha ~ nitrogen_kg_ha * variety, data = yield_data)
summary(raw_model)
```

### Residual diagnostics

```{r diagnostics-raw, fig.width=8, fig.height=4}
# Side-by-side diagnostic plots
par(mfrow = c(1, 2))

# QQ plot — checks normality of residuals
qqnorm(resid(raw_model), main = "Normal Q-Q Plot (Raw Model)")
qqline(resid(raw_model), col = "red")

# Residuals vs Fitted — checks linearity & constant variance
plot(fitted(raw_model), resid(raw_model),
     xlab = "Fitted values", ylab = "Residuals",
     main = "Residuals vs Fitted (Raw Model)")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

### 📋 Canvas Question (Section 1)

> **In 1–2 sentences: What assumption of the linear model is violated,
> and how can you tell from the diagnostic plots?**
>
> *(Submit your answer on Canvas — do not write it here.)*

------------------------------------------------------------------------

## 🛑 Stop Point 1

*Pause here — the instructor will lead a brief discussion before we
continue.*

------------------------------------------------------------------------

# Section 2 — Transformation & Re-check (~8 min)

The residual-vs-fitted plot above likely showed a **fan shape**
(variance increasing with the mean) and possibly some curvature. This is
classic **heteroscedasticity with multiplicative errors** — exactly the
situation where a **log transformation** helps.

### Apply a log transformation and re-fit

```{r transform-refit}
# Log-transform the response
yield_data$log_yield <- log(yield_data$yield_kgha)

# Re-fit the same model structure on log(yield)
log_model <- lm(log_yield ~ nitrogen_kg_ha * variety, data = yield_data)
summary(log_model)
```

### Re-run diagnostics on the transformed model

```{r diagnostics-log, fig.width=8, fig.height=4}
par(mfrow = c(1, 2))

qqnorm(resid(log_model), main = "Normal Q-Q Plot (Log Model)")
qqline(resid(log_model), col = "blue")

plot(fitted(log_model), resid(log_model),
     xlab = "Fitted values", ylab = "Residuals",
     main = "Residuals vs Fitted (Log Model)")
abline(h = 0, col = "blue", lty = 2)

par(mfrow = c(1, 1))
```

Notice that the fan shape is **reduced** and the QQ plot is closer to
the diagonal. The transformation is doing its job.

### ANOVA — Type I, Type II, and Type III

Because the design is **unbalanced** (different sample sizes per
variety) and the model includes an **interaction**, the choice of
sum-of-squares type matters. Below are all three. **Your task: pick the
correct one, delete or comment out the other two, and justify your
choice on Canvas.**

```{r anova-types}
# ------ Type I (sequential) SS ------
# Order of terms matters; results depend on which predictor is entered first.
cat("===== Type I SS (sequential) =====\n")
anova(log_model)

# ------ Type II SS ------
# Each main effect tested after the other main effect, ignoring the interaction.
cat("\n===== Type II SS =====\n")
Anova(log_model, type = "II")

# ------ Type III SS ------
# Each term tested after ALL other terms (including the interaction).
# Requires setting contrasts to sum-to-zero for valid results.
cat("\n===== Type III SS =====\n")
options(contrasts = c("contr.sum", "contr.poly"))
log_model_t3 <- lm(log_yield ~ nitrogen_kg_ha * variety, data = yield_data)
Anova(log_model_t3, type = "III")
```

### 📋 Canvas Questions (Section 2)

> 1.  **Which ANOVA type is most appropriate here, and why?** *(Hint:
>     think about the study design — is it balanced? Is the interaction
>     important?)*
>
> 2.  **Is the interaction between nitrogen and variety statistically
>     significant?** Look at the ANOVA table you chose and report the
>     p-value.
>
> *(Submit your answers on Canvas — do not write them here.)*

------------------------------------------------------------------------

## 🛑 Stop Point 2

*Pause here — the instructor will reveal the correct answer and discuss
why Type III is the right choice for an unbalanced design with a
significant interaction.*

------------------------------------------------------------------------

# Section 3 — The Wrong Model (~8 min)

Now we switch to a **completely different dataset**: seed sprouting. We
recorded whether each seed sprouted (1) or not (0), along with the
seed's size in millimeters.

**The question**: Does seed size predict the probability of sprouting?

### Load the sprouting data and fit a linear model

```{r load-data2}
# Read the sprouting dataset
sprout_data <- read.csv("data/seed_sprouting.csv")

# Quick look
str(sprout_data)
table(sprout_data$sprout)
```

```{r fit-lm-binary}
# Fit a LINEAR model to the binary outcome — this is intentionally wrong!
wrong_model <- lm(sprout ~ seed_size_mm, data = sprout_data)
summary(wrong_model)
```

### Plot: predictions from the linear model

```{r plot-wrong-model, fig.width=7, fig.height=5}
# Create a sequence of seed sizes for prediction
new_sizes <- data.frame(seed_size_mm = seq(0, 14, by = 0.1))

# Predict with 95% confidence interval
preds <- predict(wrong_model, newdata = new_sizes, interval = "confidence")
new_sizes <- cbind(new_sizes, preds)

# Plot the observed data and the linear predictions
plot(sprout_data$seed_size_mm, sprout_data$sprout,
     pch = 16, col = rgb(0, 0, 0, 0.4),
     xlab = "Seed Size (mm)", ylab = "Sprouted (0/1)",
     main = "Linear Model Predictions for Binary Outcome",
     ylim = c(-0.3, 1.5))

# Add the fitted line and confidence band
lines(new_sizes$seed_size_mm, new_sizes$fit, col = "red", lwd = 2)
lines(new_sizes$seed_size_mm, new_sizes$lwr, col = "red", lty = 2)
lines(new_sizes$seed_size_mm, new_sizes$upr, col = "red", lty = 2)

# Reference lines at 0 and 1
abline(h = 0, col = "grey50", lty = 3)
abline(h = 1, col = "grey50", lty = 3)

legend("topleft",
       legend = c("Observed", "Fitted (lm)", "95% CI"),
       pch = c(16, NA, NA), lty = c(NA, 1, 2), col = c("black", "red", "red"),
       bty = "n")
```

### Diagnostics for the wrong model

```{r diagnostics-wrong, fig.width=8, fig.height=4}
par(mfrow = c(1, 2))

qqnorm(resid(wrong_model), main = "Normal Q-Q Plot (Linear on 0/1)")
qqline(resid(wrong_model), col = "red")

plot(fitted(wrong_model), resid(wrong_model),
     xlab = "Fitted values", ylab = "Residuals",
     main = "Residuals vs Fitted (Linear on 0/1)")
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

Look at the residual plot — you should see **two distinct bands** of
residuals (one for sprout = 0, one for sprout = 1). This is a hallmark
of fitting a continuous model to binary data.

### 📋 Canvas Questions (Section 3)

> 1.  **What range do the fitted values fall in — is that a problem?**
>     *(Look at the prediction plot. Do any predicted "probabilities" go
>     below 0 or above 1?)*
>
> 2.  **What does the residual plot look like — does it meet the
>     assumptions of a linear model?** *(Describe what you see.)*
>
> 3.  **What kind of model should we use instead, and why?** *(Hint:
>     think about what family of models is designed for binary
>     outcomes.)*
>
> *(Submit your answers on Canvas — do not write them here.)*

------------------------------------------------------------------------

## 🛑 Stop Point 3 — Wrap-up

*The instructor will discuss:*

-   *Why `lm()` fails for binary outcomes*
-   *What a **generalized linear model (GLM)** is — specifically
    **logistic regression** (`glm()` with `family = binomial`)*
-   *How the logit link function keeps predicted probabilities inside
    \[0, 1\]*

We will fit GLMs properly in the next class session.
