---
title: "Functions"
format: 
  html:
    self-contained: true
---

# Writing scripts and functions

<div>

## Create a script! 📜

It is usually better to create functions (at least initially) in a script rather than a Quarto document.

Go to file \> new \> new file \> R script

Do not do this in a Quarto document

</div>

So far we have been using Quarto to turn in our assignments. Quarto (and RMarkdown) are great for reports, to present code, and sometimes to have a nice, well commented analysis.

However, more often than not, when working on our own code we write scripts in the console. That's what we will do in today's activity.

## Functions

So far, we have been using "pre-built" functions. Functions that come in the R base package, and functions that come in some packages we run. Usually the functions look the following way:

$$
\underbrace{lm}_{function}\underbrace{(formula, data) }_{input}
$$

Good news! We can create our own functions. This can be super helpful, for things we run over and over again. It is also generally better to *soft code* than to *hard code*. This allows us to do that. We can make a function that will run a predetermined script. Here is an example of a function I just made. You can copy and paste it in your r source code window and run it.

```{r}
plus10<-function(n){
  n<-n+10
  return(n)
}

plus10(7)

plus10(20)
```

Can you guess what this function does?

$$
\underbrace{plus10}_{\text{name of function}} <- function\underbrace{(n)}_{input}\text{\{} 
$$

$$
\underbrace{n<-n+10}_{code}
$$

$$
\underbrace{return(n)}_{\text{output}}\}
$$

Here is another function:

```{r}
#| eval: false

lambplot<- function(lambda1, lambda2, lambda3){
  
  
    x2 <- seq(0,qpois(0.99, max(c(lambda1,lambda2,lambda3)))+2, by = 1)
    
   selectedDatap <-c(dpois(x2, lambda1))
  
  selectedDatap2 <- c(dpois(x2, lambda2))
 
  selectedDatap3 <- c(dpois(x2, lambda3))
 

  
 df<- data.frame(y = c(selectedDatap,selectedDatap2,selectedDatap3),
                            x2 = x2,
                            lambda = factor(rep(c(paste("lambda = ", lambda1), paste("lambda = ", lambda2), paste("lambda = ", lambda3)), each = length(x2)),
                                        levels = c(paste("lambda = ", lambda1), paste("lambda = ", lambda2), paste("lambda = ", lambda3))))
  



 aplot<- ggplot(data=df, aes(x = x2+0.5, y = y, fill = lambda)) +
    geom_col(alpha = 0.75, color = "grey", linewidth = 0.1, position = "dodge") +
    scale_y_continuous("Probability density") +
    scale_x_continuous("y") +
   theme_classic()
    return(list(aplot,paste("your selections were",lambda1, lambda2, "and", lambda3)))    

  }
```

You can copy and paste it, and try to guess what it does.

Try to run it with values of 1, 5, and 10. Once you understand what it does, let's move on.

## Tasks

Now you have two tasks. These are just for class time. You can choose to do any or both

CREATE THESE TASKS IN AN R-SCRIPT. NOT A QUARTO DOCUMENT!!!

### TASK 1 🖊️

#### Goal

Write a function that simulates data for a linear relationship, fits a simple linear model to the data, and plots the results.

#### Description

Essentially, you will run the function with some inputs, and the output will be 1) a scatter plot with a line (showing the expected values; also known as trendline) with confidence interval, and 2) the model summary output. Let's make it simple and have a simple linear model with a single continuous variable as the explanatory.

#### Parts of the function:

##### Inputs (your function must take these as arguments):

-   `n`: number of observations (sample size)

-   `int`: intercept (β₀)

-   `slope`: slope (β₁)

-   `variance`: amount of random noise/variation around the line

##### Simulation setup:

-   Choose a continuous explanatory variable (x) that makes sense in your study system.

-   Simulate the x values (based on n) — a normal or uniform distribution usually works well.

-   Generate the response variable (y) using the linear equation: $$y_i = \beta_0 + \beta_1 x_i + \epsilon_i $$

-   The random error should come from a normal distribution with the variance you specify.The rnorm function asks for SD, not variance!

##### Model fitting and prediction:

-   Store the new data in a data frame (should have x and observed y)

-   Fit a simple linear model (`lm`) using your simulated data

-   Save the model summary

-   Use the `predict()` function to estimate the predicted values (aka trendline).

-   Use `interval = "prediction"` in the prediction, so you get confidence intervals around your estimates.

##### Plot:

-   Make a scatterplot of your data with the predicted line and confidence interval.

<!-- -->

-   Use `ggplot2` or base R (I recommend `ggplot2`

##### Output:

After running your function, the output should `return()` a plot and a model summary.

##### Make it complex!

If you finish this function, you can make it more complex. Allow users to also add a categorical variable, the function should work whether the user inputs a categorical variable or not.

Let users choose whether the effect is **additive** or **interactive**. Make sure the function still works even if no group is provided — in that case, it should behave like your simpler version.

### TASK 2 ✏️

#### Goal

Simulate and compare two groups with different types of count data distributions:

-   One group follows a Poisson distribution

-   The other follows a Negative Binomial distribution

#### Steps:

##### 1. Create a function

-   Simulates count data for two groups

-   Group 1 follows a **Poisson distribution**

-   Group 2 follows a **Negative Binomial distribution**

The following code should help you simulate a negative binomial distribution

```{r}
#| eval: false

dnbinom(x=0:50,size=,mu=)

```

-   `mu` = the mean (expected value)

-   `size` = the dispersion parameter ($\theta$), which controls the variance.

##### 2. Visualize the two distributions

Create a plot that shows the frequency or probability of different count values for each group. You might use `geom_bar()` or `geom_line()` depending on your preference
