---
title: "Week 1"
author: "FWF 690"
format: 
  html:
    self-contained: true
editor: visual
---

## Welcome to week 1

\importantbox{If you are reading this file before class, it may not make a lot of sense to you. That's OK! We will go through it on Wednesday}

Before we get into the weeds of data analysis, coding, and R, we should remember some basics! On Friday we will learn (or refresh) some basic R skills! If you have a lot of experience in R, just hang in there, we will get more and more complex topics as we go on! If you're brand new, also hang in there, this may seem overwhelming and very time consuming. It will get easier!

We will figure out theory and coding as we go. For all assignments, we will use Quarto. It allows you to write some neat reports while using code and plots.

### R Projects

We will learn more about projects during class. We will have our first R assignment this Friday. For the time being, just trust me, and let's do the following:

1.  Open R Studio, go to file \> new project \> new directory \> new project

2.  Name it FWF693, and use a directory that makes sense to you

3.  Download the .qmd and the .csv files IN THE NEWLY CREATED PROJECT DIRECTORY

\importantbox{If you are working on this before class, you can (and should) stop now}

### Statistics refresher

The following information will only make sense after class on Wednesday.

Open R and make sure your project is opened. Then open the .qmd file in R. We will continue there!

First of, we discussed some very important concepts during class. You can take notes here

What is statistics?

What is a population and a sample? Give examples

Data

Variable and observation

Types of data:

Categorical: Nominal and Ordinal

Numerical: Discrete and Continuous

Now, let's go back to the html file and follow the instructions to explore some different types of data.

### Data exploration

Download and read the butterfly file. Do this in a new file.

\notebox{If you are comfortable with R basics, you can write the code for this exercise directly into Quarto. To add extra code chunks, write this symbols: ```{r} }

To read data, we use the following line:

```{r}
data<-read.csv("ButterflyData.csv", stringsAsFactors = T)
```

\notebox{This only works if the file is in the current working directory. Make sure you download the file to the projects folder!}

Let's look at some of the data, the head command shows us the top 5 rows:

```{r}
head(data)
```

OK. What types of data are they? What are the variables and what are the observations?

Let's look at the structure of the file:

```{r}
str(data)
```

```{r}
data$Status<-as.factor(data$Status)
```

You can look at the whole dataset using the following:

```{r eval=F}
print(data)
```

Look at the data. Do you think there is an effect of \# of parasites on wing growth? Do you think different species have different sizes? How can we tell?

### Inferential statistics

We can use inferential statistics to answer these questions! But why? What are inferential statistics, and what is special about inferential statistics? Discuss

Thoughts:

More on that next weeks.

To understand inferential statistics, we need to understand probability.

### Plotting

Look at the data. Usually the best exploratory analysis is a visual one. We will learn about plotting during the course. We will learn about good plots, bad plots, and how to use R and ggplot to make some publication level plots.

For the purposes of this exercises, I want you to think about what type of plot you ant to make, and attempt to make it. You can use R, you can use excel, you can try to draw it. You can work in teams, but try to work mostly by yourself. We will revisit this dataset at the end of the course, and you will plot it again. Hopefully this will show some of the learning you have obtained!

### Probability

Discuss and write a definition for probability:

What is a probability distribution? Give one example

### Friday

Introduction to R, Projects, and Quarto.
