---
title: "Mixed models in glmmTMB"
format: 
  html:
    self-contained: true
---

# How to run a model in glmmTMB

## Let's run the models

Mixed effects models in glmmTMB. How do we run them?

We have both fixed and mixed effects models in glmmTMB. In this specific example, we are trying to run 4 models to figure out which random effects to use in our model selection step. The first step is to run 4 models, each with the same fixed effects, but different random effects.

Here is how you run those models in r:

$$
\underbrace{model2}_{\text{Model Name}}<- \underbrace{glmmTMB}_{\text{function for mixed effects models}} (\underbrace{\text{Richness~NAP*Exposure}}_{\text{Fixed effects}} \underbrace{+ (1|Beach)}_{Random effects}, data = rikz, \underbrace{family = poisson}_{Distribution})
$$ {#eq-model2}

In our example we are LEAVING FIXED EFFECTS THE SAME. THEY ARE ALWAYS Richness\~Nap\*Exposure.

However, we are exploring 4 random effects:

1.  No random effects
2.  Random intercept (model 2, the one showed above)
3.  Random slope
4.  Random intercept and random slope

Let's run each model.

### Model 1:

The no random effects model is:

$$
log(ERichness) \sim \beta_0 + \beta_1 (NAP)_i + \beta_2 Exposure2_i ... 
$$

$$
Richness_i \sim Poisson(ERichness)_i
$$

And the code to run it is:

```{r}
#| eval: false

model1<-glmmTMB(Richness~NAP*Exposure, data = rikz, family = poisson)
```

It has no random effect... so, it's just run as a simple glm.

### Model 2

The random intercept model is:

$$ log(ERichness) \sim (\beta_0 + \gamma_j) + \beta_1 (NAP)_i + \beta_2 Exposure2_i ...  $$

$$ Richness_i \sim Poisson(ERichness)_i $$

And the way to introduce than randomness $\gamma_j$ is:

$$
\underbrace{model2}_{\text{Model Name}}<- \underbrace{glmmTMB}_{\text{function for mixed effects models}} (\underbrace{\text{Richness~NAP*Exposure}}_{\text{Fixed effects}} \underbrace{+ (1|Beach)}_{Random effects}, data = rikz, \underbrace{family = poisson}_{Distribution})
$$

or:

```{r}
#| eval: false

model2<-glmmTMB(Richness~NAP*Exposure + (1|Beach), data = rikz, family = poisson)

```

If you see the equation, we need to estimate one gamma for each beach. That is why we use (1\|Beach).

### Model 3

Random slope:

$$ log(ERichness) \sim \beta_0 + (\beta_1 * \psi_j * NAP_i )+ \beta_2 Exposure2_i ...  $$

$$ Richness_i \sim Poisson(ERichness)_i $$

Here, there is a random slope ($\psi_j$), and it affects NAP. Exposure is categorical, so, it doesn't hav a slope. This is how we run it:

$$
\underbrace{model3}_{\text{Model Name}}<- \underbrace{glmmTMB}_{\text{function for mixed effects models}} (\underbrace{\text{Richness~NAP*Exposure}}_{\text{Fixed effects}} \underbrace{+ (0+NAP|Beach)}_{Random effects}, data = rikz, \underbrace{family = poisson}_{Distribution})
$$

Here, we don't want it to estimate random intercepts, hence the 0. However, we want it to estimate random slopes, so, that's why we ass NAP

```{r}
#| eval: false

model3<-glmmTMB(Richness~NAP*Exposure + (0+NAP|Beach), data = rikz, family = poisson)
```

### Model 4:

Random slope and intercept:

$$ log(ERichness) \sim (\beta_0 + \gamma_j) + (\beta_1 * \psi_j * NAP_i )+ \beta_2 Exposure2_i ...  $$

$$ Richness_i \sim Poisson(ERichness)_i $$

Here, there is a random slope ($\psi_j$) and random intercept ($\gamma_i$). This is how we run it:

$$
\underbrace{model4}_{\text{Model Name}}<- \underbrace{glmmTMB}_{\text{function for mixed effects models}} (\underbrace{\text{Richness~NAP*Exposure}}_{\text{Fixed effects}} \underbrace{+ (1+NAP|Beach)}_{Random effects}, data = rikz, \underbrace{family = poisson}_{Distribution})
$$

Or

```{r}
#| eval: false


model4<-glmmTMB(Richness~NAP*Exposure + (1+NAP|Beach), data = rikz, family = poisson)
```
