---
title: "Poisson Counts, Log Transformations & Intro to Poisson GLMs"
author: "PLSC/ANSC/SNR 571"
format: html
editor: visual
execute:
  output: false
---

```{r setup, include=FALSE}
library(car)
library(ggplot2)
```

For this activity you will need the `car` package and the `ggplot2` package.

## Overview

This in-class coding activity has four sections. All code is provided: your job is to run the code, look at the output, and answer the discussion questions on Canvas.

I will start with a hummingbird feeder dataset. The response variable is the number of feeder visits during a fixed observation period. Run each chunk in order, read the short notes, and then submit your answers on Canvas.

------------------------------------------------------------------------

# Section 1. Load the data and fit the wrong model (~7 min)

I will start simple and only use `sucrose_pct`, which is the strongest continuous predictor in this dataset.

### Load the data

```{r load-data}
poisson_data <- read.csv("poisson_dataset.csv")

poisson_data$feeder_color <- as.factor(poisson_data$feeder_color)
poisson_data$site <- as.factor(poisson_data$site)

str(poisson_data)
summary(poisson_data)
```

### Fit a linear model to the raw counts

```{r fit-wrong-model}
wrong_model <- lm(visits ~ sucrose_pct, data = poisson_data)
summary(wrong_model)
```

### Residual diagnostics

```{r diagnostics-wrong, fig.width=8, fig.height=4}
par(mfrow = c(1, 2))

qqnorm(resid(wrong_model), main = "Normal Q-Q Plot (Raw Counts)")
qqline(resid(wrong_model), col = "red")

plot(
  fitted(wrong_model), resid(wrong_model),
  xlab = "Fitted values", ylab = "Residuals",
  main = "Residuals vs Fitted (Raw Counts)"
)
abline(h = 0, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

Let's also look at the raw relationship with a straight line added.

```{r plot-wrong-model}
ggplot(poisson_data, aes(x = sucrose_pct, y = visits)) +
  geom_point(alpha = 0.7, color = "darkgreen") +
  stat_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "black") +
  theme_classic() +
  labs(
    x = "Sucrose concentration (%)",
    y = "Hummingbird visits"
  )
```

### 📋 Canvas Question (Section 1)

> **In 1-2 sentences: Which linear-model assumptions look violated here, and what do you see in the plots that makes you say that?**
>
> *(Submit your answer on Canvas, do not write it here.)*

------------------------------------------------------------------------

## 🛑 Stop Point 1

*Pause here: discussion*

------------------------------------------------------------------------

# Section 2. Log transform and re-check (~8 min)

Because some feeders have zero visits, I will use `log(visits + 1)` instead of `log(visits)`.

### Transform the response and plot the relationship

```{r transform-response}
poisson_data$log_visits <- log(poisson_data$visits + 1)

ggplot(poisson_data, aes(x = sucrose_pct, y = log_visits)) +
  geom_point(alpha = 0.7, color = "darkblue") +
  stat_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "black") +
  theme_classic() +
  labs(
    x = "Sucrose concentration (%)",
    y = "log(visits + 1)"
  )
```

### Fit a linear model to the transformed response

```{r fit-log-lm}
log_model <- lm(log_visits ~ sucrose_pct, data = poisson_data)
summary(log_model)
```

```{r diagnostics-log, fig.width=8, fig.height=4}
par(mfrow = c(1, 2))

qqnorm(resid(log_model), main = "Normal Q-Q Plot (Log Response)")
qqline(resid(log_model), col = "blue")

plot(
  fitted(log_model), resid(log_model),
  xlab = "Fitted values", ylab = "Residuals",
  main = "Residuals vs Fitted (Log Response)"
)
abline(h = 0, col = "blue", lty = 2)

par(mfrow = c(1, 1))
```

The log transformation often helps because the relationship looks closer to linear and the spread is more stable. A Poisson GLM also uses a log idea, but it models the mean count with a log link while keeping a Poisson distribution for the response on the original count scale.

### 📋 Canvas Question (Section 2)

> **In 1-2 sentences: What is similar between the transformed `lm()` and a Poisson `glm()`, and what is different about what each model assumes for the response?**
>
> *(Submit your answer on Canvas, do not write it here.)*

------------------------------------------------------------------------

## 🛑 Stop Point 2

*Pause here*

------------------------------------------------------------------------

# Section 3. Fit the Poisson GLM (~8 min)

Now I will fit the model that matches the count response.

```{r fit-poisson-glm}
poisson_model <- glm(
  visits ~ sucrose_pct,
  family = poisson(link = "log"),
  data = poisson_data
)

summary(poisson_model)
```

The slope in this model is on the log scale. If `sucrose_pct` increases by one unit, the expected log mean count changes by the slope estimate. On the original scale, the expected mean count is multiplied by `exp(beta1)` for a one-unit increase in sucrose concentration.

### Compare the transformed `lm()` and the Poisson `glm()`

```{r compare-models}
coef(summary(log_model))
coef(summary(poisson_model))
```

### ANOVA

With one predictor, the test is straightforward. I will use Type II here to keep the workflow consistent with the `car` package tools we will use later.

```{r poisson-anova}
Anova(poisson_model, type = "II")
```

### 📋 Canvas Question (Section 3)

> **After looking at both summaries, what differences do you notice between the transformed `lm()` and the Poisson `glm()`? Think about the coefficients, standard errors, and how you would interpret fitted values.**
>
> *(Submit your answer on Canvas, do not write it here.)*

### Prediction and plotting on the original count scale

```{r poisson-predictions}
new_data <- data.frame(
  sucrose_pct = seq(
    min(poisson_data$sucrose_pct),
    max(poisson_data$sucrose_pct),
    length.out = 200
  )
)

pred_link <- predict(poisson_model, newdata = new_data, type = "link", se.fit = TRUE)

new_data$fit <- exp(pred_link$fit)
new_data$lwr <- exp(pred_link$fit - 1.96 * pred_link$se.fit)
new_data$upr <- exp(pred_link$fit + 1.96 * pred_link$se.fit)
```

```{r plot-poisson-glm, fig.width=7, fig.height=5}
ggplot(poisson_data, aes(x = sucrose_pct, y = visits)) +
  geom_point(alpha = 0.65, color = "darkgreen") +
  geom_ribbon(
    data = new_data,
    aes(y = fit, ymin = lwr, ymax = upr),
    inherit.aes = FALSE,
    alpha = 0.2,
    fill = "skyblue3"
  ) +
  geom_line(
    data = new_data,
    aes(y = fit),
    inherit.aes = FALSE,
    color = "navy",
    linewidth = 1
  ) +
  theme_classic() +
  labs(
    x = "Sucrose concentration (%)",
    y = "Expected hummingbird visits"
  )
```

------------------------------------------------------------------------

## 🛑 Stop Point 3

*Pause here: discussion*

------------------------------------------------------------------------

# Section 4. Looking ahead to next week (~5 min)

So far I only used one predictor, but this dataset has more information. Besides `sucrose_pct`, you also have `feeder_height_cm`, `shade_pct`, `feeder_color`, and `site`.

### 📋 Canvas Question (Section 4)

> **If you wanted to explain hummingbird visits more fully, what would you do next? What model would you run, and how could you compare different biological hypotheses or competing models?**
>
> *(Submit your answer on Canvas, do not write it here.)*

We will cover multiple predictors and model comparison next week.
