---
title: "Polynomial Models: Wrong Fits, Diagnostics, and Model Comparison"
author: "PLSC/ANSC/SNR 571"
format: html
editor: visual
execute:
  output: false
---

## Packages needed

Make sure you have these packages installed and loaded

```{r setup}
library(car)
library(ggplot2)
library(AICcmodavg)
```

For this activity you will need the `car` package, the `ggplot2` package, and the `AICcmodavg` package.

## Overview

This in-class coding activity has four sections: a wrong linear model for a plant dataset, a quadratic correction, a simple Poisson model for count data, and a polynomial Poisson comparison.

Most of the code is provided. Your job is to run the chunks in order, look at the output, and answer the discussion questions on Canvas.

I want you to pay attention to one big idea all day: a model can still be linear in the coefficients even when it contains polynomial terms like `x`, `I(x^2)`, and `I(x^3)`.

------------------------------------------------------------------------

# Section 1. Plant biomass: fit the wrong model first (\~10 min)

I will start with a greenhouse dataset. Plants were grown under different fertilizer rates, and the response variable is aboveground biomass in grams.

The biology suggests that more fertilizer should help at first, but not forever. Still, I want to start with the wrong model on purpose.

### Load the data

```{r load-biomass-data}
plant_data <- read.csv("fertilizer_biomass_quadratic.csv")

str(plant_data)
summary(plant_data)
```

### Fit a linear model to the raw relationship

```{r fit-biomass-linear, output = FALSE}
biomass_linear <- lm(biomass_g ~ fertilizer_g_m2, data = plant_data)
summary(biomass_linear)
```

### Residual diagnostics

```{r diagnostics-biomass-linear, fig.width=8, fig.height=4, output = FALSE}
par(mfrow = c(1, 2))

qqnorm(resid(biomass_linear), main = "Normal Q-Q Plot (Linear Biomass Model)")
qqline(resid(biomass_linear), col = "red")

plot(
  fitted(biomass_linear), resid(biomass_linear),
  xlab = "Fitted values", ylab = "Residuals",
  main = "Residuals vs Fitted (Linear Biomass Model)"
)
abline(h = mean(resid(biomass_linear)), col = "red", lty = 2)

fv <- fitted(biomass_linear)
rv <- resid(biomass_linear)
sm <- loess(rv ~ fv)
newx <- seq(min(fv), max(fv), length.out = 200)
lines(newx, predict(sm, newdata = newx), col = "blue", lwd = 2)

legend(
  "topright",
  legend = c("Residual mean", "Loess smooth"),
  col = c("red", "blue"),
  lty = c(2, 1),
  lwd = c(1, 2),
  bty = "n"
)

par(mfrow = c(1, 1))
```

### Plot the raw relationship with the wrong fit

```{r plot-biomass-linear, fig.width=7, fig.height=5, output = FALSE}
ggplot(plant_data, aes(x = fertilizer_g_m2, y = biomass_g)) +
  geom_point(alpha = 0.75, color = "darkgreen") +
  stat_smooth(method = "lm", formula = y ~ x, se = TRUE, color = "black") +
  theme_classic() +
  labs(
    x = "Fertilizer rate (g/m2)",
    y = "Plant biomass (g)"
  )
```

### 📋 Canvas Question (Section 1)

> **In 1-2 sentences: What do the residuals-vs-fitted plot and the model plot suggest about the linearity assumption? What pattern tells you the linear model is wrong?**
>
> *(Submit your answer on Canvas, do not write it here.)*

------------------------------------------------------------------------

## 🛑 Stop Point 1

*Pause here: discussion*

------------------------------------------------------------------------

# Section 2. Fit the quadratic model and compare models (\~15 min)

Now I will fit the model that matches the biological story more closely.

### Fit a quadratic model

```{r fit-biomass-quadratic, output = FALSE}
biomass_quadratic <- lm(
  biomass_g ~ fertilizer_g_m2 + I(fertilizer_g_m2^2),
  data = plant_data
)
summary(biomass_quadratic)
```

### Re-check diagnostics

```{r diagnostics-biomass-quadratic, fig.width=8, fig.height=4, output = FALSE}
par(mfrow = c(1, 2))

qqnorm(resid(biomass_quadratic), main = "Normal Q-Q Plot (Quadratic Biomass Model)")
qqline(resid(biomass_quadratic), col = "blue")

plot(
  fitted(biomass_quadratic), resid(biomass_quadratic),
  xlab = "Fitted values", ylab = "Residuals",
  main = "Residuals vs Fitted (Quadratic Biomass Model)"
)
abline(h = mean(resid(biomass_quadratic)), col = "blue", lty = 2)

fv2 <- fitted(biomass_quadratic)
rv2 <- resid(biomass_quadratic)
sm2 <- loess(rv2 ~ fv2)
newx2 <- seq(min(fv2), max(fv2), length.out = 200)
lines(newx2, predict(sm2, newdata = newx2), col = "darkgreen", lwd = 2)

legend(
  "topright",
  legend = c("Residual mean", "Loess smooth"),
  col = c("blue", "darkgreen"),
  lty = c(2, 1),
  lwd = c(1, 2),
  bty = "n"
)

par(mfrow = c(1, 1))
```

### Compare the linear fit and the quadratic fit

```{r plot-biomass-both, fig.width=7, fig.height=5, output = FALSE}
new_biomass <- data.frame(
  fertilizer_g_m2 = seq(
    min(plant_data$fertilizer_g_m2),
    max(plant_data$fertilizer_g_m2),
    length.out = 200
  )
)

new_biomass$linear_fit <- predict(biomass_linear, newdata = new_biomass)
new_biomass$quadratic_fit <- predict(biomass_quadratic, newdata = new_biomass)

ggplot(plant_data, aes(x = fertilizer_g_m2, y = biomass_g)) +
  geom_point(alpha = 0.7, color = "darkgreen") +
  geom_line(
    data = new_biomass,
    aes(x = fertilizer_g_m2, y = linear_fit, color = "Linear model"),
    linewidth = 1
  ) +
  geom_line(
    data = new_biomass,
    aes(x = fertilizer_g_m2, y = quadratic_fit, color = "Quadratic model"),
    linewidth = 1
  ) +
  scale_color_manual(values = c("Linear model" = "black", "Quadratic model" = "blue")) +
  theme_classic() +
  labs(
    x = "Fertilizer rate (g/m2)",
    y = "Plant biomass (g)",
    color = "Model"
  )
```

::: callout-note
## AIC vs AICc

AIC is the original information criterion and is most comfortable in large samples.

AICc is a corrected version of AIC that adds a small-sample penalty when the sample size is not very large relative to the number of parameters.

As sample size gets larger, AICc gets closer and closer to AIC.

A simple rule of thumb is this: if your dataset is modest in size, or if you are not sure, prefer AICc. In practice I usually use AICc for classroom datasets like this one.
:::

### Model comparison with `AICcmodavg`

F `aictab()` reports AICc and orders models from best to worst.

```{r biomass-aicc, output = FALSE}
aictab(
  cand.set = list(linear = biomass_linear, quadratic = biomass_quadratic)
)
```

### 📋 Canvas Question (Section 2)

> **After fitting the quadratic model, what changed in the residual pattern? Which model is better supported by the information-criterion tables, and why?**
>
> *(Submit your answer on Canvas, do not write it here.)*

------------------------------------------------------------------------

## 🛑 Stop Point 2

*Pause here*

------------------------------------------------------------------------

# Section 3. Seedling counts: start with the simple Poisson model (\~10 min)

Now we will switch to count data from an ecology study. Field crews counted oak seedlings across sites with different canopy cover.

I will start with the simplest Poisson GLM even though I already suspect the relationship is more complicated than a straight line on the link scale.

### Load the data

```{r load-seedling-data, output = FALSE}
seedling_data <- read.csv("seedling_counts_cubic.csv")

str(seedling_data)
summary(seedling_data)
```

### Fit a Poisson GLM with only a linear term

```{r fit-seedling-linear, output = FALSE}
seedling_linear <- glm(
  oak_seedlings ~ canopy_cover_pct,
  family = poisson(link = "log"),
  data = seedling_data
)
summary(seedling_linear)
```

### Plot the simple Poisson fit

```{r plot-seedling-linear, fig.width=7, fig.height=5, output = FALSE}
new_seedlings <- data.frame(
  canopy_cover_pct = seq(
    min(seedling_data$canopy_cover_pct),
    max(seedling_data$canopy_cover_pct),
    length.out = 200
  )
)

pred_linear <- predict(seedling_linear, newdata = new_seedlings, type = "link", se.fit = TRUE)
new_seedlings$linear_fit <- exp(pred_linear$fit)
new_seedlings$linear_lwr <- exp(pred_linear$fit - 1.96 * pred_linear$se.fit)
new_seedlings$linear_upr <- exp(pred_linear$fit + 1.96 * pred_linear$se.fit)

ggplot(seedling_data, aes(x = canopy_cover_pct, y = oak_seedlings)) +
  geom_point(alpha = 0.7, color = "darkgreen") +
  geom_ribbon(
    data = new_seedlings,
    aes(x = canopy_cover_pct, y = linear_fit, ymin = linear_lwr, ymax = linear_upr),
    inherit.aes = FALSE,
    alpha = 0.2,
    fill = "skyblue3"
  ) +
  geom_line(
    data = new_seedlings,
    aes(x = canopy_cover_pct, y = linear_fit),
    inherit.aes = FALSE,
    color = "navy",
    linewidth = 1
  ) +
  theme_classic() +
  labs(
    x = "Canopy cover (%)",
    y = "Oak seedling count"
  )
```

### Pearson residuals for the simple Poisson model

```{r diagnostics-seedling-linear, fig.width=7, fig.height=5, output = FALSE}
plot(
  fitted(seedling_linear), resid(seedling_linear, type = "pearson"),
  xlab = "Fitted values", ylab = "Pearson residuals",
  main = "Pearson Residuals vs Fitted (Linear Poisson Model)"
)
abline(h = mean(resid(seedling_linear, type = "pearson")), col = "red", lty = 2)

fv3 <- fitted(seedling_linear)
rv3 <- resid(seedling_linear, type = "pearson")
sm3 <- loess(rv3 ~ fv3)
newx3 <- seq(min(fv3), max(fv3), length.out = 200)
lines(newx3, predict(sm3, newdata = newx3), col = "blue", lwd = 2)
```

### 📋 Canvas Question (Section 3)

> **Does the simple Poisson model seem flexible enough here? What do the raw plot and the Pearson residual plot suggest about the shape of the relationship?**
>
> *(Submit your answer on Canvas, do not write it here.)*

------------------------------------------------------------------------

## 🛑 Stop Point 3

*Pause here: discussion*

------------------------------------------------------------------------

# Section 4. Add polynomial terms to the Poisson GLM and compare models (\~15 min)

Now I will fit models of increasing complexity and compare them.

```{r fit-seedling-polynomials, output = FALSE}
seedling_quadratic <- glm(
  oak_seedlings ~ canopy_cover_pct + I(canopy_cover_pct^2),
  family = poisson(link = "log"),
  data = seedling_data
)

seedling_cubic <- glm(
  oak_seedlings ~ canopy_cover_pct + I(canopy_cover_pct^2) + I(canopy_cover_pct^3),
  family = poisson(link = "log"),
  data = seedling_data
)

summary(seedling_quadratic)
summary(seedling_cubic)
```

### Compare the candidate models with `AICcmodavg`

```{r seedling-aicc, output = TRUE}
seedling_tab <- aictab(
  cand.set = list(
    linear = seedling_linear,
    quadratic = seedling_quadratic,
    cubic = seedling_cubic
  )
)

seedling_tab
```

Look carefully at the delta values and the Akaike weights. A delta near 0 means the model is strongly supported relative to the others in the table. The weights can be read like relative support across the candidate set.

### Plot the linear and cubic fits on the original count scale

```{r plot-seedling-both, fig.width=7, fig.height=5}
pred_cubic <- predict(seedling_cubic, newdata = new_seedlings, type = "link", se.fit = TRUE)
new_seedlings$cubic_fit <- exp(pred_cubic$fit)

ggplot(seedling_data, aes(x = canopy_cover_pct, y = oak_seedlings)) +
  geom_point(alpha = 0.7, color = "darkgreen") +
  geom_line(
    data = new_seedlings,
    aes(x = canopy_cover_pct, y = linear_fit, color = "Linear Poisson"),
    linewidth = 1
  ) +
  geom_line(
    data = new_seedlings,
    aes(x = canopy_cover_pct, y = cubic_fit, color = "Cubic Poisson"),
    linewidth = 1
  ) +
  scale_color_manual(values = c("Linear Poisson" = "black", "Cubic Poisson" = "blue")) +
  theme_classic() +
  labs(
    x = "Canopy cover (%)",
    y = "Expected oak seedling count",
    color = "Model"
  )
```

### Check the cubic model residual pattern

```{r diagnostics-seedling-cubic, fig.width=7, fig.height=5}
plot(
  fitted(seedling_cubic), resid(seedling_cubic, type = "pearson"),
  xlab = "Fitted values", ylab = "Pearson residuals",
  main = "Pearson Residuals vs Fitted (Cubic Poisson Model)"
)
abline(h = mean(resid(seedling_cubic, type = "pearson")), col = "blue", lty = 2)

fv4 <- fitted(seedling_cubic)
rv4 <- resid(seedling_cubic, type = "pearson")
sm4 <- loess(rv4 ~ fv4)
newx4 <- seq(min(fv4), max(fv4), length.out = 200)
lines(newx4, predict(sm4, newdata = newx4), col = "darkgreen", lwd = 2)
```

### 📋 Canvas Question (Section 4)

> **Which Poisson model is best supported by the AICc table? Use the delta values and Akaike weights in your answer, and explain what the cubic terms add biologically.**
>
> *(Submit your answer on Canvas, do not write it here.)*

------------------------------------------------------------------------

