---
title: "Polynomial Models Lecture Companion"
author: "PLSC/ANSC/SNR 571"
format: html
editor: visual
execute:
  output: false
---

## Overview

This is an instructor-facing outline for two 50 minute class meetings that line up with `polynomial_models.qmd` and the datasets `fertilizer_biomass_quadratic.csv` and `seedling_counts_cubic.csv`.

Use this as a live teaching guide. The goal is not to make slides. The goal is to know what to write, what to show, where to pause, and what code to type.

------------------------------------------------------------------------

# Lecture 1. Quadratic structure in a linear model (about 50 min)

## 0 to 8 min. Open with the big idea

- Write: `y = beta0 + beta1 x`
- Then write: `y = beta0 + beta1 x + beta2 x^2`
- Say clearly that both models are linear in the coefficients.
- Ask: what changes when I add `x^2`? What shape can the model now make?
- Show the raw fertilizer and biomass scatterplot before fitting anything.

## 8 to 18 min. Live code the wrong model first

What to type:

```{r eval=FALSE}
library(ggplot2)
plant_data <- read.csv("fertilizer_biomass_quadratic.csv")
biomass_linear <- lm(biomass_g ~ fertilizer_g_m2, data = plant_data)
summary(biomass_linear)
```

What to highlight:

- Identify the response and predictor.
- Remind students that a significant slope does not guarantee the model shape is correct.
- Point out that the fitted line is trying to average over curvature.

Concept checks to ask:

- What is the response variable?
- What is the explanatory variable?
- If the biology has an optimum, does a straight line make sense across the full range?

## 18 to 30 min. Diagnostics for the wrong model

What to type:

```{r eval=FALSE}
par(mfrow = c(1, 2))
qqnorm(resid(biomass_linear))
qqline(resid(biomass_linear), col = "red")
plot(fitted(biomass_linear), resid(biomass_linear))
abline(h = mean(resid(biomass_linear)), col = "red", lty = 2)
```

What to show:

- The residuals vs fitted plot is the key plot here.
- Point at the curved pattern and say: the mean structure is wrong.
- Mention that the residual mean line is near 0, but that does not rescue the model.
- If time allows, add a loess line and ask students what it is tracking.

Pause and ask:

- Which assumption is the main problem here?
- If the residuals bend, what does that tell us about the model equation?

## 30 to 40 min. Fit the quadratic model

What to type:

```{r eval=FALSE}
biomass_quadratic <- lm(
  biomass_g ~ fertilizer_g_m2 + I(fertilizer_g_m2^2),
  data = plant_data
)
summary(biomass_quadratic)
```

What to write on the board:

- `E(y) = beta0 + beta1 x + beta2 x^2`
- If `beta2 < 0`, the curve opens downward.
- The turning point happens at `-beta1 / (2 * beta2)`.

What to highlight:

- The model is still a linear model.
- We changed the shape by changing the predictors, not by changing the estimation framework.

## 40 to 50 min. Compare fits and introduce AIC vs AICc

What to type:

```{r eval=FALSE}
library(AICcmodavg)
aictab(
  cand.set = list(linear = biomass_linear, quadratic = biomass_quadratic),
  second.ord = FALSE
)
aictab(
  cand.set = list(linear = biomass_linear, quadratic = biomass_quadratic)
)
```

What to say:

- AIC is the original criterion and is aimed at larger samples.
- AICc adds a small sample correction.
- As sample size grows, AICc approaches AIC.
- Rule of thumb: if you are unsure, prefer AICc.

Concept checks:

- Does the best supported model need the highest R-squared?
- What does a lower information criterion mean?

------------------------------------------------------------------------

# Lecture 2. Cubic structure in a Poisson GLM (about 50 min)

## 0 to 8 min. Connect back to last class

- Review the five-step workflow from the assignment.
- Write the Poisson GLM form: `log(lambda) = beta0 + beta1 x`
- Then extend it: `log(lambda) = beta0 + beta1 x + beta2 x^2 + beta3 x^3`
- Say that the polynomial is on the link scale.

## 8 to 18 min. Load the count dataset and fit the simple Poisson model

What to type:

```{r eval=FALSE}
seedling_data <- read.csv("seedling_counts_cubic.csv")
seedling_linear <- glm(
  oak_seedlings ~ canopy_cover_pct,
  family = poisson(link = "log"),
  data = seedling_data
)
summary(seedling_linear)
```

What to show:

- Raw counts against canopy cover.
- The fitted mean curve from the simple Poisson model.
- Remind students that the slope is on the log scale.

Concept checks:

- What is the response variable here?
- Why is `lm()` not the first choice for these data?
- If the link is log, what function gets us back to the count scale?

## 18 to 28 min. Diagnose the simple Poisson fit

What to type:

```{r eval=FALSE}
plot(
  fitted(seedling_linear),
  resid(seedling_linear, type = "pearson")
)
abline(h = mean(resid(seedling_linear, type = "pearson")), col = "red", lty = 2)
```

What to highlight:

- Focus on the shape of the residual pattern, not on perfect randomness.
- If the linear predictor is too simple, the residual smooth should still show structure.
- Explain why that suggests missing polynomial terms.

Pause and ask:

- What does the residual pattern suggest we should try next?
- Is this a distribution problem, a mean-structure problem, or both?

## 28 to 40 min. Fit quadratic and cubic Poisson models

What to type:

```{r eval=FALSE}
seedling_quadratic <- glm(
  oak_seedlings ~ canopy_cover_pct + I(canopy_cover_pct^2),
  family = poisson(link = "log"),
  data = seedling_data
)

seedling_cubic <- glm(
  oak_seedlings ~ canopy_cover_pct + I(canopy_cover_pct^2) + I(canopy_cover_pct^3),
  family = poisson(link = "log"),
  data = seedling_data
)
```

What to write on the board:

- `log(lambda) = beta0 + beta1 x + beta2 x^2 + beta3 x^3`
- `lambda = exp(beta0 + beta1 x + beta2 x^2 + beta3 x^3)`

What to highlight:

- The cubic model can capture changes in curvature that the linear model cannot.
- The interpretation of each single coefficient is harder, so plotting predicted values becomes especially important.

## 40 to 50 min. Model comparison and wrap-up

What to type:

```{r eval=FALSE}
seedling_tab <- aictab(
  cand.set = list(
    linear = seedling_linear,
    quadratic = seedling_quadratic,
    cubic = seedling_cubic
  )
)
seedling_tab
```

What to show:

- The delta AICc column.
- The Akaike weights.
- A plot with the linear and cubic fitted curves on the count scale.

What to ask:

- Which model is best supported?
- If the cubic model wins, is that because it is more complex or because it captures a real pattern?
- How would you explain the pattern biologically in one or two sentences?

## End-of-day take home message

- Polynomial terms change model shape without changing the basic regression framework.
- Diagnostics tell us when a model is too simple.
- Information criteria help compare candidate models that represent different biological ideas.
- For small or moderate classroom datasets, AICc is usually the safer default.
