---
title: "Simulation parameters"
format: revealjs
---

## Questions about reading

What are the simulation parameters?

$$
P(\# \text{ Heads} = 2) \approx \frac{\text{Number of times \{# Heads = 2\} } }{\text{# Experiments}}
$$

## Questions about reading

What are the simulation parameters?

$$ P(\text{Event}) \approx \frac{\text{Number of times  { # Event} } }{\text{# Experiments}} $$

## In R

Number of times you get 2 heads when tossing 3 coins.

First step:

Toss 1 coin 3 times and record the number of heads:

```{r echo=TRUE}
rbinom(1,3,0.5)
```

## In R

Second step:

Toss 1 coin 3 times and record the number of heads. Let's repeat the experiment 1,000 times

```{r echo=TRUE}
rbinom(1000,3,0.5)
```

## In R

Third step:

Count how many of those times we got 2 heads from N experiments and obtain the P

$$
P(\# \text{ Heads} = 2) \approx \frac{\text{Number of times \{# Heads = 2\} } }{\text{# Experiments}}
$$

```{r echo=TRUE}
n<-1000
sum(rbinom(n,3,0.5)==2)/n
```

## Probability of winning the lottery once if I play it 1,000,000 times

```{r echo=TRUE}
n<-1000000
sum(rbinom(n,10000,3.42e-09)>0)/n
```

## Main issue...

We usually don't know the probability of a success :(

## In R

$$
P(\# \text{ Heads} = 2) \approx \frac{\text{Number of times \{# Heads = 2\} } }{\text{# Experiments}}
$$

```{r echo=TRUE}
n<-1000
sum(rbinom(n,3,0.5)==2)/n
```
