---
title: "Last assignment (optional)"
author: "Dr. Alejandro Molina Moctezuma"
format:
  html: 
    embed-resources: true
    fig_height: 4.5
    fig_width: 4.5
    highlight: tango
    number_sections: true
    toc: true
    toc_depth: 2
    theme: cosmo
    fontsize: 1.1em
    linestretch: 1.7
editor: visual
knitr:
  opts_chunk: 
    collapse: true
    header: "Last Assignment"
execute: 
  warning: false
  tidy: true
---

```{r echo=FALSE}
library(tidyverse)
```

## Repeated measures

As discussed in class, repeated measures can be on of the most complex experimental designs to analyze. There are many reasons for this, but the main one is that time is hard to model, and that there is temporal autocorrelation.

## Chicken data

We will do the same example as in class. Make sure to look at the slides! Download the chickdata file and read it

```{r}
chickdata<- read.csv("chickdata.csv")
```

As a reminder, this is what the data looks like:

```{r echo=FALSE}
library(kableExtra)
chickdata_wide <- tidyr::pivot_wider(chickdata, names_from = Time,
                                     names_prefix = "Day", values_from = Weight)
chickdata_wide[c(1,2,10, 11 ,19, 20, 28, 29),] %>%
  kable(format = "html", col.names = c("Chick", "Diet", "Day 0", "Day 2", "Day 4", "Day 6", "Day 8", "Day 10", "Day 12", "Day 14", "Day 16", "Day 18")) %>%
  add_header_above(c(" " = 1, " " = 1, "Weight" = 10))
```

The data that you downloaded is in the long format. What you see in the table is the same data in wide format. We can use the `tidyr` function `pivot_longer` and `pivot_wider` to go from wide to long and long to wide. This can be super useful.

Most analyses in R require our data to be in the long format.

This is some of our data in graphical form:

```{r echo=FALSE}
ggplot(chickdata[chickdata$Chick %in% c(1,2,10, 11 ,19, 20, 28, 29),], aes(x = Time, y = Weight, color = as.factor(Diet), linetype = as.factor(Chick))) +
  geom_path(lwd=1.3) +
  geom_point(size=3) +
  scale_x_continuous("Day") +
  scale_y_continuous("Weight") +
  guides(linetype = "none")+
  theme_classic()
```

### Additive model (split plot)

In the additive model, we can use the following equation:

$$
y_{ijk} = \mu + \alpha_i + \beta_j + \alpha\beta_{ij} + \delta_{ik} + \epsilon_{ijk}
$$

where,

$\mu$ is the grand mean, $\alpha_i$ is the effect of diet, $\beta_j$ is the effect of time, $\alpha \beta_{ij}$ is the interaction of time and diet, $\delta_ik$ is the effect of the kth subject receiving the ith treatment, and $\epsilon_{ijk}$ is the residual error.

To fit this model, we use the followinf:

```{r}
modelaov<-aov(Weight ~ Diet * Time + Error(Chick),data=chickdata)
summary(modelaov)
```

While this might be ok, there is an issue. Our errors are not independent, so we might be underestimating our p-value. This is because of an an issue called temporal autocorrelation. How much a chick grew in time 2 will affect its growth in time 4, and that growth will affect time 6 and so on. We need to take this into account by adjusting our p-values up. You should ONLY do the following, if the original ANOVA was significant. If it wasn't, there is no need, as the p-value will only go up.

In order to adjust our p-values, we estimate the sphericity of the data. There are two methods to do this, but the good news is that R runs both.

First off, we need "wide" data:

```{r}
chickdata_wide <- tidyr::pivot_wider(chickdata, names_from = Time,
                                     names_prefix = "Day", values_from = Weight)
```

Check the new dataset.

To run a spherical test, we run a `manova` :

```{r}
man<-manova(cbind(Day0, Day2, Day4, Day6, Day8,Day10,Day12,Day14,Day16,Day18) ~ Diet, data = chickdata_wide)
```

This model is already taking into account both the effects of diet, time and the interactions. However, it is running it a bit differently. The results from this are:

```{r}
anova(man, X = ~1, test = "Spherical")
```

Where Intercept is the effect of time, and Diet is the interaction of time and diet.

In this case, our p-values for the interaction went from p\<0.0001 to p\>0.05, therefore they are not significant anymore. This is a result of the p-value adjustment.

### Profile analysis

Probably the better way to do it is with a profile analysis. In this case, rather than looking at the wight of the chickens, we look at the weight difference between measurements.

To do this, we do the following:

```{r}
man2<-manova(cbind(Day2-Day0,Day4- Day2, Day6-Day4, Day8-Day6,Day10- Day8,Day12-Day10,Day14-Day12,Day16-Day14,Day18-Day16) ~ Diet, data = chickdata_wide)
```

See how we have only 9 intervals, despite having 10 times.

```{r}
summary.aov(man2)
```

To plot these results, we do the following:

```{r}
Weightm<-chickdata_wide[,3:12]
weightdif<-Weightm[2:10]-Weightm[1:9]
colnames(weightdif) <- paste("interval", 1:9, sep=".")
kable(head(weightdif))
```

This is the data that we have now:

```{r echo=FALSE}
chickdata_widex <- tidyr::pivot_wider(chickdata, names_from = Time,
                                     names_prefix = "Time", values_from = Weight)

chickdata_wide2 <- dplyr::mutate(chickdata_widex, int1 = Time2 - Time0,
                                 int2 = Time4 - Time2,
                                 int3 = Time6 - Time4,
                                 int4 = Time8 - Time6,
                                 int5 = Time10 - Time8,
                                 int6 = Time12 - Time10,
                                 int7 = Time14 - Time12,
                                 int8 = Time16 - Time14,
                                 int9 = Time18 - Time16)

chickdata2 <- tidyr::pivot_longer(chickdata_wide2,
                                  cols = c(int1, int2, int3, int4, int5, int6, int7, int8, int9),
                                  names_to = "Interval",
                                  values_to = "delta")
chickdata2$Interval <- rep(1:9, 36)
ggplot(chickdata2[chickdata2$Chick %in% c(1,2,10, 11 ,19, 20, 28, 29),],
       aes(x = Interval, y = delta, color = as.factor(Diet), linetype = as.factor(Chick))) +
  geom_path(lwd=1.7) +
  geom_point(size=3) +
  scale_x_continuous("Interval", breaks = c(1, 3, 5, 7, 9)) +
  scale_y_continuous("Weight change") +
  guides(linetype = "none")+
  theme_classic()

```

This is a complex analysis, so, it's OK if you get a bit lost.

We need to estimate the means for each week, for each diet:

```{r}
Diet1<-colMeans(weightdif[1:9,])
Diet2<-colMeans(weightdif[10:18,])
Diet3<-colMeans(weightdif[19:27,])
Diet4<-colMeans(weightdif[28:36,])
```

Now, we calculate the Standard Errors:

```{r}
SE <- sqrt(diag(stats:::vcov.mlm(man2)))
SE <- SE[names(SE)==":(Intercept)"] # Only use "intercept" SEs
unname(SE) ## Ignore the names
```

Finally, we can put everything together, and plot:

```{r}
growthDF  <- data.frame(interval = rep(1:9, 4),
                        Diet = rep(c("1", "2", "3", "4"), each = 9),
                        growth = c(Diet1, Diet2, Diet3, Diet4), 
                        SE = rep(SE, 2))


ggplot(growthDF, aes(x = interval, y = growth, color = Diet)) +
  geom_path() +
  geom_point() +
  geom_errorbar(aes(ymin = growth - SE, ymax = growth + SE), width = 0.1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_continuous("Time interval") +
  scale_y_continuous("Growth rate", limits  = c(-1, 40))+
  theme_classic()
  
```

## Mixed model with correlation

Finally, you can run it as a mixed model with correlated data:

```{r}
library(nlme)
lmm2 <- lme(fixed = Weight ~ Diet *Time,
            random =  ~ (Time-1)|Chick,
            data = chickdata,
            correlation = corAR1())
summary(lmm2)
anova(lmm2)
```

This is what I would recommend. And it actually shows an effect!

## Power analysis

This is a very straightforward analysis.

There are four parameters in a model that are dependent on each other:

1.  Degrees of freedom
2.  Effect (how big of a difference?)
3.  $\alpha$ generally 0.05
4.  Power

If you have one, you can estimate the rest. For effect, you can choose an effect size metric or you can also give exact differences (and standard deviation)

For example, if we are comparing the effects of fertilizer on plant growth. We believe a growth of 8 cm would be biologically important, we want an $\alpha$ of 0.05 and a power of 0.5 (generally pretty good power). We have observed these populations have a standard deviation of \~3 we can do the following:

```{r}
power.t.test(delta=5,power=.80,sd=3)
```

Essentially, we only need 7 plants in each group!

If we want to find:

1\) smaller differences we need more plants:

```{r}
power.t.test(delta=2,power=.80,sd=3)
```

37 plants per group to be able to find these super small differences!

### Power analysis ANOVA

In Anova, we are able to use the package `pwr`. Because in ANOVA, oftentimes we have complex models with multiple factors, it is hard to completely provide an effect size in absolute terms. Instead of this, we can use Cohen's conventional effect size.

For our example, we will do a 2 factorial effect. The Cohen's value would be:

```{r}
library(pwr)
eff<-cohen.ES("f2", size = "medium")$effect.size 
eff
```

For a small value it would be:

```{r}
eff<-cohen.ES("f2", size = "small")$effect.size 
eff
```

In this case, I will use 0.9, which is a value between both of them.

Let's assume in this case we are doing a full factorial 3 by 3 ANOVA

```{r}
pwr.f2.test(u = 1, # I'm assuming you plan to test a 1 df effect
            f2 = 0.09, # We're using Cohen's effect size guidelines, shown above
            sig.level = .05, # Our alpha
            power = .80) # Our desired power
```

In this case the degrees of freedom of the "error term" are 87.17. If we add the 8 df's from the 3\*3 design plus the intercept, we get that we need 97 observations TOTAL. In this case 11 observations per group would be enough.
