---
title: "Overdispersion part 1"
format:
  revealjs:
    incremental: true   
    scrollable: true
    chalkboard: true
server: shiny
---

```{r}
library(ggplot2)
library(dplyr)
library(car)
library(readr)
```

## Next classes

\~ 6 weeks left

1.  Zero-inflated model and lienar models (lab: zero inflated and mixed effects) LONG
2.  Git and shiny: Linear models reviews (lab: repeated measures and mixed effects)
3.  Last four weeks: GAMs and multivariate (ordination) –\> Readings and long-ish labs

**Last project**

## Count data

Remember the Poisson distribution?

-   count data = Poisson

-   But usually, Poisson is not the best distribution to test count data

-   Two reasons: over-inflation and too many zeroes

## Poisson distribution

```{r}
#| panel: fill
fluidRow(
column(8,  plotOutput(outputId = "plot2", 
             height = "500px")),

  column(4,
numericInput("nump", "lambda", 0.1, min=0, max=50),
numericInput("nump2", "lambda", 10, min=0, max=50),
numericInput("nump3", "lambda", 20, min=0, max=50)))



```

```{r}
#| context: server

#| fig-width: 8
#| fig-height: 5

library(ggplot2)
x2 <- seq(0, 50, by = 1)

selectedDatap <- reactive({
    c(dpois(x2, input$nump))
  })
selectedDatap2 <- reactive({
    c(dpois(x2, input$nump2))
  })
selectedDatap3 <- reactive({
    c(dpois(x2, input$nump3))
  })



df3 <- reactive({data.frame(y = c(selectedDatap(),selectedDatap2(),selectedDatap3()),
                 x2 = x2,
                 mu = factor(rep(c("lambda = 0.1", "lambda = 1", "lambda = 5"), each = length(x2)),
                                 levels = c("lambda = 0.1", "lambda = 1", "lambda = 5")))
  
})

output$plot2<-renderPlot({
ggplot(data=df3(), aes(x = x2+0.5, y = y, fill = mu)) +
  geom_col(alpha = 0.75, color = "grey", linewidth = 0.1, position = "dodge") +
  scale_y_continuous("Probability density") +
  scale_x_continuous("y") +
    theme(legend.position="none")
})

```

## Poisson distribution

In poisson we have the following:

-   $$
    log(\lambda_i) = \beta_0 + \beta_1x_{1,i)} + ... 
    $$

-   $$
    y_i \sim Poisson (\lambda_i)
    $$

-   $y_i$ is the response

-   $\lambda_i$ is the expected value

-   $\lambda_i$ is also the variance

## Normal distribution (lm)

-   $$
    E[y_i] \sim \beta_0 + \beta_1x_{1,i} + ... 
    $$

-   $$
    y_i \sim Normal(E[y_i],\sigma^2)
    $$

## Plot again

```{r}

#| echo: false
#| fig-width: 8
#| fig-height: 5

fluidRow(
column(8,  plotOutput(outputId = "plot1", 
             height = "500px")),

  column(4,
numericInput("num", "mu1", 0.1, min=-40, max=40),

numericInput("sd", "sd1", 1, min=-40, max=40),

numericInput("num2", "mu2", 10, min=-40, max=40),
numericInput("sd2", "sd2", 1, min=-40, max=40),

numericInput("num3", "mu3", 10, min=-40, max=40),
numericInput("sd3", "sd3", 10, min=-40, max=40)))
```

```{r}
#| context: server

library(ggplot2)

selectedData <- reactive({
    c(dnorm(x, input$num,input$sd))
  })
selectedData2 <- reactive({
    c(dnorm(x, input$num2,input$sd2))
  })
selectedData3 <- reactive({
    c(dnorm(x, input$num3,input$sd3))
  })


x <- seq(-50, 50, by = .1)
df <- reactive ({data.frame(y = c(selectedData() , selectedData2(), selectedData3()),
                 x = x,
                 mu = factor(rep(c("mu = 0.1", "mu = 1", "mu = 3"), each = length(x)),
                                 levels = c("mu = 0.1", "mu = 1", "mu = 3")))
})

output$plot1<- renderPlot({
  
ggplot(data=df(), aes(x = x, y = y, color = mu, fill=mu)) +
  geom_col(alpha = 0.95, linewidth = 0.1, position = "dodge") +
  scale_y_continuous("Probability density") +
  scale_x_continuous("y") +
  scale_color_discrete(labels = c(expression(paste(mu, "= 0.1, ", sigma, "= 1")), 
                                 expression(paste(lambda, "= 1")), 
                                 expression(paste(lambda, "= 3")))) +
    scale_fill_discrete(labels = c(expression(paste(mu, "= 0.1, ", sigma, "= 1")), 
                                 expression(paste(lambda, "= 1")), 
                                 expression(paste(lambda, "= 3")))) +
    theme(legend.position="none")
  
})
```

## Plot

```{r}
plotOutput(outputId = "plot5", height = "600px")
```

```{r}
#| context: server

  x5 <- runif(100, 0, 60)
  ey <- 10 + 1*x5 
  eps <- rnorm(100, 0, 3)
  y5 <- ey + eps
  df5 <- data.frame(x5, y5)
  lm_fit <- lm(y5 ~ x5, data = df5)
  
  k <- 2.5
  sigma <- sigma(lm_fit)
  ab <- coef(lm_fit); a <- ab[1]; b <- ab[2]
  
  x5 <- seq(-k*sigma, k*sigma, length.out = 50)
  y5 <- dnorm(x5, 0, sigma)/dnorm(0, 0, sigma) * 3
  
  x0 <- 5
  y0 <- a+b*x0
  path1 <- data.frame(x = y5 + x0, y = x5 + y0)
  segment1 <- data.frame(x = x0, y = y0 - k*sigma, xend = x0, yend = y0 + k*sigma)
  x0 <- 20
  y0 <- a+b*x0 
  path2 <- data.frame(x = y5 + x0, y = x5 + y0)
  segment2 <- data.frame(x = x0, y = y0 - k*sigma, xend = x0, yend = y0 + k*sigma)
  x0 <- 35
  y0 <- a+b*x0 
  path3 <- data.frame(x = y5 + x0, y = x5 + y0)
  segment3 <- data.frame(x = x0, y = y0 - k*sigma, xend = x0, yend = y0 + k*sigma)
  x0 <- 50
  y0 <- a+b*x0 
  path4 <- data.frame(x = y5 + x0, y = x5 + y0)
  segment4 <- data.frame(x = x0, y = y0 - k*sigma, xend = x0, yend = y0 + k*sigma)

  output$plot5<- renderPlot({
  ggplot(df5, mapping = aes(x=x5, y=y5)) + 
    geom_point(color="#446E9B", alpha = 0.5) + 
    geom_smooth(method='lm', se=FALSE, color="#D47500", alpha = 0.5) + 
    geom_path(aes(x,y), data = path1, linewidth = 1, color = "#3CB521") + 
    geom_path(aes(x,y), data = path2, linewidth = 1, color = "#3CB521") + 
    geom_path(aes(x,y), data = path3, linewidth = 1, color = "#3CB521") +
    geom_path(aes(x,y), data = path4, linewidth = 1, color = "#3CB521") +
    geom_segment(aes(x=x,y=y,xend=xend,yend=yend), data = segment1, linewidth = 0.6, 
                 color = "#CD0200") +
    geom_segment(aes(x=x,y=y,xend=xend,yend=yend), data = segment2, linewidth = 0.6,
                 color = "#CD0200") +
    geom_segment(aes(x=x,y=y,xend=xend,yend=yend), data = segment3, linewidth = 0.6,
                 color = "#CD0200") +
    geom_segment(aes(x=x,y=y,xend=xend,yend=yend), data = segment4, linewidth = 0.6,
                 color = "#CD0200") +
    theme_classic()
  })
```

## Poisson distribution

In poisson we have the following:

-   $$ log(\lambda_i) = \beta_0 + \beta_1x_{1,i)} + ...  $$

-   $$ y_i \sim Poisson (\lambda_i) $$

-   $y_i$ is the response

-   $\lambda_i$ is the expected value


